Signal2Image Modules in Deep Neural Networks for EEG Classification

### 2021-09-21 ms.tex
init

\documentclass[conference]{IEEEtran}
\usepackage{multirow}
\usepackage{siunitx}
\usepackage{diagbox}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{tikz}
\renewcommand\theadfont{\bfseries}
\IEEEoverridecommandlockouts

@@

One common approach that previous studies have used for classifying EEG signals was feature extraction from the frequency and time-frequency domains utilizing the theory behind EEG band frequencies~\cite{langkvist2012sleep}: delta (0.5-4 Hz), theta (4-8 Hz), alpha (8-13 Hz), beta (13-20 Hz) and gamma (20-64 Hz).
Truong et al.~\cite{truong2018convolutional} used Short-Time Fourier Transform (STFT) on a 30 second sliding window to train a three layer CNN on stacked time-frequency representations for seizure prediction and evaluated their method on three EEG databases.
Khan et al.~\cite{khan2018focal} transformed the EEGs in time-frequency domain using multi-scale wavelets and then trained a six layer CNN on these stacked multi-scale representations for predicting the focal onset seizure demonstrating promising results.

@@

\section{DATA}
The UCI EEG epileptic seizure recognition dataset~\cite{andrzejak2001indications} consists of $500$ signals each one with $4097$ samples (23.5 seconds).
The dataset is annotated into five classes with $100$ signals for each class (in parenthesis the shortened label names used in Fig. \ref{fig:highleveloverview} and \ref{fig:signal2imageoutputs}):
\begin{enumerate}
	\item healthy patient while having his eyes open (Open),
	\item healthy patient while having his eyes closed (Closed),
	\item patient with tumor taken from healthy area (Healthy),
	\item patient with tumor taken from tumor area (Tumor),
	\item patient while having seizure activity (Epilepsy)
\end{enumerate}

@@

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[]
		\node[] at (-4.5, 0){\includegraphics[scale=0.2,angle=90]{tmp/signal-epilepsy}};
		\node[align=left] at (-4.2, 0) {$x_i$};
		\draw[dashed,->] (-4, 0) -- (-3.8, 0);
		\node[draw, minimum height=2.55cm] at (-3.5, 0) {$m$};
		\draw[dashed,->] (-3.2, 0) -- (-3, 0);
		\node[] at (-1.7, 0){\includegraphics[scale=0.41,angle=90]{tmp/cnn-epilepsy}};
		\draw[dashed,->] (-0.4, 0) -- (-0.2, 0);
		\node[draw, minimum height=2.55cm] at (0.1, 0) {$b_{d}$};
		\draw[dashed,->] (0.4, 0) -- (0.6, 0);
		\node[align=left] at (0.8, 0) {$\hat{y_i}$};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.67, 1.1) {\footnotesize Open 0.1\%};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.74, 0.55) {\footnotesize Closed 0.2\%};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.78, 0) {\footnotesize Healthy 0.9\%};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.8, -0.55) {\footnotesize Tumor 34.7\%};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.9, -1.1) {\footnotesize Epilepsy 64.1\%};
	\end{tikzpicture}
	\caption{High level overview of a feed-forward pass of the combined methods.
	$x_i$ is the input, $m$ is the Signal2Image module, $b_{d}$ is the 1D or 2D architecture `base model' for $d=1,2$ respectively and $\hat{y_i}$ is the predicted output.
	The names of the labels are depicted at the right along with the predictions for this example signal.
	The image between $m$ and $b_{d}$ depicts the output of the one layer CNN Signal2Image module, while the `signal as image' and spectrogram have intermediate images as those depicted at the second and third row of Fig.~\ref{fig:signal2imageoutputs}.
	Arrows denote the flow of the feed-forward pass.
	For the 1D architectures $m$ is omitted and no intermediate image is generated.}
	\label{fig:highleveloverview}
\end{figure}

\section{METHODS}
\subsection{Definitions}
We define the dataset $D=\{x_i, y_i\}_{i=1...N}$ where $x_i \in \mathbb{Z}^n$ and $y_i \in \{1, 2, 3, 4, 5\}$ denote the $i^{th}$ input signal with dimensionality $n=178$ and the $i^{th}$ label with five possible classes respectively.
$N=11500$ is the number of observations.

@@

We finally define the combinations of $m$ and $b_d$ as the members $c$ of the set of combined models $C$.
Using the previous definitions, the aim of this paper is the evaluation of the set of models $C$, where $C$ is the combined set of $M$ and $B$ i.e. $C=M\times B$ w.r.t. time performance and class accuracy trained on $D$.

@@

\begin{figure*}[!t]
	\centering
	\rotatebox[origin=c]{90}{Original Signal}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-eyes-open}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-eyes-closed}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-healthy-area}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-tumor-area}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-epilepsy}
	\end{subfigure}

	\rotatebox[origin=c]{90}{`Signal as Image'}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-eyes-open}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-eyes-closed}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-healthy-area}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-tumor-area}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-epilepsy}
	\end{subfigure}

	\rotatebox[origin=c]{90}{Spectrogram}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-eyes-open}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-eyes-closed}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-healthy-area}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-tumor-area}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-epilepsy}
	\end{subfigure}

	\rotatebox[origin=c]{90}{\hspace{5mm} one layer CNN}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-eyes-open}
		\caption{Open}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-eyes-closed}
		\caption{Closed}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-healthy-area}
		\caption{Healthy}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-tumor-area}
		\caption{Tumor}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-epilepsy}
		\caption{Epilepsy}
	\end{subfigure}
	\caption{Visualizations of the original signals and the outputs of the S2Is for each class.
	The x, y-axis of the first row are in \SI{}{\micro V} and time samples respectively.
	The x, y-axis of the rest of the subfigures denote spatial information, since we do not inform the `base model' the concept of time along the x-axis or the concept of frequency along the y-axis.
	Higher pixel intensity denotes higher amplitude.}
	\label{fig:signal2imageoutputs}
\end{figure*}

@@

Out of the 11500 signals we used $76\%$, $12\%$ and $12\%$ of the data ($8740,1380,1380$ signals) as training, validation and test data respectively.
No artifact handling or preprocessing was performed.
All networks were trained for $100$ epochs and model selection was performed using the best validation accuracy out of all the epochs.
We used PyTorch~\cite{paszke2017automatic} for implementing the neural network architectures and training/preprocessing was done using a NVIDIA Titan X Pascal GPU 12GB RAM and a 12 Core Intel i7-8700 CPU @ 3.20GHz on a Linux operating system.

\begin{table*}[!t]
	\caption{Test accuracies (\%) for combined models.
	The second row indicates the number of layers.
	Bold indicates the best accuracy for each base model.}
	\label{table:results}
	\begin{minipage}{\textwidth}
		\setlength\tabcolsep{0.6pt}
		\centering
		\input{tmp/results.tex}
	\end{minipage}
\end{table*}

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{multirow}
\usepackage{siunitx}
\usepackage{diagbox}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{adjustbox}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{tikz}
\renewcommand\theadfont{\bfseries}
\IEEEoverridecommandlockouts{}

@@

One common approach that previous studies have used for classifying EEG signals was feature extraction from the frequency and time-frequency domains utilizing the theory behind EEG band frequencies~\cite{langkvist2012sleep}: delta (0.5--4 Hz), theta (4--8 Hz), alpha (8--13 Hz), beta (13--20 Hz) and gamma (20--64 Hz).
Truong et al.~\cite{truong2018convolutional} used Short-Time Fourier Transform (STFT) on a 30 second sliding window to train a three layer CNN on stacked time-frequency representations for seizure prediction and evaluated their method on three EEG databases.
Khan et al.~\cite{khan2018focal} transformed the EEGs in time-frequency domain using multi-scale wavelets and then trained a six layer CNN on these stacked multi-scale representations for predicting the focal onset seizure demonstrating promising results.

@@

\section{DATA}
The UCI EEG epileptic seizure recognition dataset~\cite{andrzejak2001indications} consists of $500$ signals each one with $4097$ samples (23.5 seconds).
The dataset is annotated into five classes with $100$ signals for each class (in parenthesis the shortened label names used in Fig.~\ref{fig:highleveloverview} and~\ref{fig:signal2imageoutputs}):
\begin{enumerate}
	\item healthy patient while having his eyes open (Open),
	\item healthy patient while having his eyes closed (Closed),
	\item patient with tumor taken from healthy area (Healthy),
	\item patient with tumor taken from tumor area (Tumor),
	\item patient while having seizure activity (Epilepsy)
\end{enumerate}

@@

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[]
		\node[] at (-4.5, 0){\includegraphics[scale=0.2,angle=90]{tmp/signal-epilepsy.png}};
		\node[align=left] at (-4.2, 0) {$x_i$};
		\draw[dashed,->] (-4, 0) -- (-3.8, 0);
		\node[draw, minimum height=2.55cm] at (-3.5, 0) {$m$};
		\draw[dashed,->] (-3.2, 0) -- (-3, 0);
		\node[] at (-1.7, 0){\includegraphics[scale=0.41,angle=90]{tmp/cnn-epilepsy.png}};
		\draw[dashed,->] (-0.4, 0) -- (-0.2, 0);
		\node[draw, minimum height=2.55cm] at (0.1, 0) {$b_{d}$};
		\draw[dashed,->] (0.4, 0) -- (0.6, 0);
		\node[align=left] at (0.8, 0) {$\hat{y_i}$};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.67, 1.1) {\footnotesize Open 0.1\%};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.74, 0.55) {\footnotesize Closed 0.2\%};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.78, 0) {\footnotesize Healthy 0.9\%};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.8, -0.55) {\footnotesize Tumor 34.7\%};
		\node[minimum width=0.5cm, minimum height=0.5cm] at (1.9, -1.1) {\footnotesize Epilepsy 64.1\%};
	\end{tikzpicture}
	\caption{High level overview of a feed-forward pass of the combined methods.
	$x_i$ is the input, $m$ is the Signal2Image module, $b_{d}$ is the 1D or 2D architecture `base model' for $d=1,2$ respectively and $\hat{y_i}$ is the predicted output.
	The names of the labels are depicted at the right along with the predictions for this example signal.
	The image between $m$ and $b_{d}$ depicts the output of the one layer CNN Signal2Image module, while the `signal as image' and spectrogram have intermediate images as those depicted at the second and third row of Fig.~\ref{fig:signal2imageoutputs}.
	Arrows denote the flow of the feed-forward pass.
	For the 1D architectures $m$ is omitted and no intermediate image is generated.}\label{fig:highleveloverview}
\end{figure}

\section{METHODS}
\subsection{Definitions}
We define the dataset $D=\{x_i, y_i\}_{i=1\ldots N}$ where $x_i \in \mathbb{Z}^n$ and $y_i \in \{1, 2, 3, 4, 5\}$ denote the $i^{th}$ input signal with dimensionality $n=178$ and the $i^{th}$ label with five possible classes respectively.
$N=11500$ is the number of observations.

@@

We finally define the combinations of $m$ and $b_d$ as the members $c$ of the set of combined models $C$.
Using the previous definitions, the aim of this paper is the evaluation of the set of models $C$, where $C$ is the combined set of $M$ and $B$ i.e. $C=M\times B$ w.r.t.\ time performance and class accuracy trained on $D$.

@@

\begin{figure*}[!t]
	\centering
	\rotatebox[origin=c]{90}{Original Signal}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-eyes-open.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-eyes-closed.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-healthy-area.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-tumor-area.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.2]{tmp/signal-epilepsy.png}
	\end{subfigure}

	\rotatebox[origin=c]{90}{`Signal as Image'}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-eyes-open.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-eyes-closed.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-healthy-area.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-tumor-area.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/signal-as-image-epilepsy.png}
	\end{subfigure}

	\rotatebox[origin=c]{90}{Spectrogram}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-eyes-open.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-eyes-closed.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-healthy-area.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-tumor-area.png}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/spectrogram-epilepsy.png}
	\end{subfigure}

	\rotatebox[origin=c]{90}{\hspace{5mm} one layer CNN}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-eyes-open.png}
		\caption{Open}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-eyes-closed.png}
		\caption{Closed}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-healthy-area.png}
		\caption{Healthy}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-tumor-area.png}
		\caption{Tumor}
	\end{subfigure}
	\begin{subfigure}{0.19\linewidth}
		\centering
		\includegraphics[scale=0.5]{tmp/cnn-epilepsy.png}
		\caption{Epilepsy}
	\end{subfigure}
	\caption{Visualizations of the original signals and the outputs of the S2Is for each class.
	The x, y-axis of the first row are in \SI{}{\micro{} V} and time samples respectively.
	The x, y-axis of the rest of the subfigures denote spatial information, since we do not inform the `base model' the concept of time along the x-axis or the concept of frequency along the y-axis.
	Higher pixel intensity denotes higher amplitude.}\label{fig:signal2imageoutputs}
\end{figure*}

@@

Out of the 11500 signals we used $76\%$, $12\%$ and $12\%$ of the data ($8740,1380,1380$ signals) as training, validation and test data respectively.
No artifact handling or preprocessing was performed.
All networks were trained for $100$ epochs and model selection was performed using the best validation accuracy out of all the epochs.
We used PyTorch~\cite{paszke2017automatic} for implementing the neural network architectures and training/preprocessing was done using a NVIDIA Titan X Pascal GPU 12GB RAM and a 12 Core Intel i7--8700 CPU @ 3.20GHz on a Linux operating system.

\begin{table*}[!t]
	\caption{Test accuracies (\%) for combined models.
	The second row indicates the number of layers.
	Bold indicates the best accuracy for each base model.}\label{table:results}
	\begin{adjustbox}{width=\textwidth}
		\input{tmp/results.tex}
	\end{adjustbox}
\end{table*}
